---
title: "Bayesian Logistic Regression"
author: "Ella Chee"
date: "2026-02-27"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Bayesian Model for Wildfire Detection

## Get data from Kaggle
```{r, echo=TRUE}
library(kaggler)

# Authenticate creds
#kgl_auth(username = "ellachee21", key = "KGAT_107e527829b7a3f5505e98b50142d7db")

# Download a dataset
# system("/Users/nutellachee/anaconda3/bin/kaggle datasets download -d abdelghaniaaba/wildfire-prediction-dataset --unzip -p ./data")
```


## Load data

```{r}
library(dplyr)
library(tidyverse)

# Helper to load both classes (no wildfire/wildfire) for given folder
load_data <- function(folder_name) {

  nowildfire_list <- list.files(paste0("data/", folder_name, "/nowildfire"), full.names = TRUE)
  wildfire_list <- list.files(paste0("data/", folder_name, "/wildfire"), full.names = TRUE)
  
  print(sprintf("%d images in no wildfire class", length(nowildfire_list)))
  print(sprintf("%d images in wildfire class", length(wildfire_list)))
  
  return(list(nowildfire_list, wildfire_list))
  
}

```

Load validation, training, testing sets

```{r}
# Load validation data
print("VALIDATION:")
valid_results <- load_data("valid")
valid_nowildfire <- valid_results[[1]]
valid_wildfire <- valid_results[[2]]

# Load training data
print("TRAINING:")
train_results <- load_data("train")
train_nowildfire <- train_results[[1]]
train_wildfire <- train_results[[2]]

# Load testing data
print("TESTING:")
test_results <- load_data("test")
test_nowildfire <- test_results[[1]]
test_wildfire <- test_results[[2]]


```

## Preview image data

```{r, echo=TRUE}
library(magick)

# Randomly select image
set.seed(42)
random_img <- sample(test_wildfire, 1)

# Load image into R
img <- image_read(random_img)
print(img)

# Load image as array of values
img_array <- as.numeric(image_data(img, channels = "rgb"))
dim(img_array)

```
## Data preparation

```{r}
library(magick)
library(dplyr)

# Using each pixel as a separate feature
load_images_to_df <- function(image_paths, label, batch_size = 500, resize = 35) {
  
  # Batch to increase efficiency
  batches <- split(image_paths, ceiling(seq_along(image_paths) / batch_size))
  results <- lapply(batches, function(batch) {
    
    imgs <- tryCatch(
      image_read(batch) |>
        image_resize(paste0(resize, "x", resize, "!")),
      error = function(e) { message("Batch error: ", e$message); NULL }
    )
    if (is.null(imgs)) return(NULL)
    
    n <- length(batch)
    p <- resize * resize
    
    # Pre-allocate matrix
    X <- matrix(NA_real_, nrow = n, ncol = p)
    
    for (i in seq_len(n)) {
      arr <- image_data(imgs[i], channels = "rgb") # [channel, x, y]
      
      R <- as.integer(arr[1,,]) # Extract separate channels
      G <- as.integer(arr[2,,])
      B <- as.integer(arr[3,,])
      
      pixel_vals <- R * 256^2 + G * 256 + B # Combine into single value per pixel
      X[i, ] <- as.numeric(pixel_vals) # Flatten to vector
    }
    image_destroy(imgs)
    
    # Convert to dataframe
    df <- as.data.frame(X)
    colnames(df) <- paste0("px_", seq_len(p))
    #df$name <- batch
    df$label <- label
    df
  })
  bind_rows(results)
}

# Load each dataset
dfs <- list()

for (name in c("train", "valid", "test")) {
  nowildfire_df <- load_images_to_df(get(paste0(name, "_nowildfire")), 0, 500, 16)
  wildfire_df <- load_images_to_df(get(paste0(name, "_wildfire")), 1, 500, 16)
  
  # Combine and shuffle
  df <- bind_rows(nowildfire_df, wildfire_df)
  set.seed(42)
  shuffled_df <- df[sample(1:nrow(df)), ]
  
  dfs[[name]] <- shuffled_df
  print(sprintf("Loaded %s dataset", name))
}

# Apply to each dataset
train_df <- dfs[["train"]]
valid_df <- dfs[["valid"]]
test_df <- dfs[["test"]]

```


## Model set-up

1. Resized 350x350px to 35x35px --> flattened to 1225 px values = 1225 features,
with chains = 4, 
iter = 20, 
warmup = 3, 
68 divergent chains after warmup, 
largest R-hat is Inf, indicating chains have not mixed, 
vector memory limit of 24.0 GB reached

2. Resized to 16x16 px --> flattened to 256 px values = 256 features, 
with chains = 4, 
iter = 100, 
warmup = 18, 
327 divergent transitions after warmup,
largest R-hat is 30.33, indicating chains have not mixed
vector memory limit of 24.0 Gb reached, see mem.maxVSize()

3. Resized to 16x16 px --> flattened to 256 px values = 256 features, 
with chains = 4, 
iter = 500, 
warmup = 90, 
616 divergent transitions after warmup,
largest R-hat is R-hat is 7.38, indicating chains have not mixed
vector memory limit of 24.0 Gb reached, see mem.maxVSize()

4. Resized to 16x16 px --> flattened to 256 px values = 256 features, 
with chains = 4, 
iter = 1000, 
warmup = 180, 
199 divergent transitions after warmup,
largest R-hat is R-hat is 4.75, indicating chains have not mixed

4. Resized to 16x16 px --> flattened to 256 px values = 256 features, 
with chains = 6, 
iter = 10000, 
warmup = 1800, 
2540 divergent transitions after warmup,
largest R-hat is R-hat is 5.48, indicating chains have not mixed

```{r}
library(brms)

# Set default prior (use Bernoulli for binary classification)
default_prior <- default_prior(label ~ .,
              data = train_df,
              family = bernoulli("logit"))

# Creating Bayesian logistic regression model, use all pixels as features
brm_logit <- brm(label ~ .,
                family = bernoulli("logit"),
                data = train_df,
                chains = 6,
                iter = 10000,
                warmup = 1800,
                prior = default_prior)

# Get summary stats on posterior
summary(brm_logit)

post_draws <- as.data.frame(brm_logit)
```


## Model validation

```{r}
library(caret)

# Check posterior predictive fit
pp_check(brm_logit)

# Plot individual features
plot(brm_logit, variable = c("b_px_1", "b_px_128", "b_px_256"))

# Evaluate on validation set
pred_probs <- posterior_epred(brm_logit, newdata = valid_df)
pred_label <- as.numeric(colMeans(pred_probs) > 0.5)

# Confusion matrix
confusionMatrix(factor(pred_label, levels=c(1,0)), 
                factor(valid_df$label, levels=c(1,0)), 
                mode = "everything")

```


## Model testing
```{r}
# Evaluate on test set
pred_probs <- posterior_epred(brm_logit, newdata = test_df)
pred_label <- as.numeric(colMeans(pred_probs) > 0.5)

# Confusion matrix, get F1, precision, recall
test_conf_matrix <- confusionMatrix(factor(pred_label, levels=c(1,0)), 
                factor(test_df$label, levels=c(1,0)), 
                mode = "everything")
test_conf_matrix

# Get per-class accuracies
test_conf_mat_2 <- table(Predicted = pred_label, Actual = test_df$label)
true_positives <- diag(test_conf_mat_2) # TP
actual_counts <- colSums(test_conf_mat_2) # Total per class
per_class_accuracy <- true_positives / actual_counts # Per-class acc
print(per_class_accuracy)

# Generate confusion matrix figure
library(ggplot2)
plt <- as.data.frame(test_conf_matrix$table)
ggplot(plt, aes(Prediction, Reference, fill = Freq)) +
  geom_tile() + 
  geom_text(aes(label = Freq)) +
  scale_fill_gradient(low = "steelblue", high = "white") +
  ggtitle("Bayesian LR Confusion Matrix") +
  theme_minimal()

```
